---
title: "(1) Supervised Learning Models"
format: html
---

**Mention that you can just copy the code and insert it as an Rmd!**

Supervised learning models can make predictions after seeing lots of data with the correct answers and then discovering the connections between the elements in the data that produce the correct answers. This is like a student learning new material by studying old exams that contain both questions and answers. Once the student has trained on enough old exams, the student is well prepared to take a new exam. *These ML systems are "supervised" in the sense that a human gives the ML system data with the known correct results*

Packages needed for this session:

```{r}
#| message: false

library(mlbench) #where PimaIndiansDiabetes database is found
library(tidyverse) #for minimal data wrangling 
library(ggplot2) #already part of tidyverse - so a bit redundant
library(tidymodels) #for modeling

theme_set(theme_bw()) # to help in plot visualization (white background)
```

Load diabetes dataset (already available by installing package [mlbench](https://mlbench.github.io)). As described yesterday, this is a very popular dataset and examples of its use can be found here: [machine learning examples](https://www.kaggle.com/datasets/mathchi/diabetes-data-set/code)

```{r}
data("PimaIndiansDiabetes")
```

In the environment now you should see PimaIndiansDiabetes dataframe loaded

Lets now select only two of this columns `age` and `glucose` and store it as a new dataframe

```{r}
Data <- PimaIndiansDiabetes %>%
        select(age, glucose)
```

We have *768* observations/rows, so lets cut it down to just 80, for the sake of easier visualization and take a look

```{r}
DataSmaller <- Data[1:80,]

head(DataSmaller)

```

Now lets visualize this relationship

```{r}
#| fig-align: 'center'
#| fig-width: 6
#| fig-height: 4

ggplot(DataSmaller, aes(x = age, y = glucose)) + 
geom_point() + 
labs(x = "Age (Feature)", y = "Glucose (Label)")

```

## A Supervised Learning Algorithm: Linear Regression

Suppose we want to predict our glucose level based on our age with this dataset. We could fit a learning algorithm to our dataset and create a predictive model. In this case the *learning algorithm* is going to be *linear regression*. ggplot already contains the functionality of fitting and visualizing this linear regression model through `geom_smooth` as seen below (this is the solution from the practical from the first day):

```{r}
#| fig-align: 'center'
#| fig-width: 6
#| fig-height: 4

ggplot(DataSmaller, aes(x = age, y = glucose)) + 
geom_point() + 
labs(x = "Age (Feature)", y = "Glucose (Label)") +  
geom_smooth(method='lm', se = FALSE)

```

This amounts to using the `lm` function to fit the data as you have seen in the Statistics Module [statistics recap](https://anasrana.github.io/ems-practicals/linear-regression.html)

To recap, the linear regression algorithm is as follows:

In ML, we write the equation as

$$
y = \beta_0 + \beta_1\, x_1 
$$

where

-   y = outcome label ( column to predict )
-   \beta\_0 = sometimes known as bias, it is the intercept of the line and is a *parameter* of the model
-   \beta\_1 = weight of the feature/column/x 1 - same as slope in equation of a line (if we only had 2 dimensions) and is a *parameter* of the model
-   x_1 = feature/column 1/ input data

And how do we find out the two parameters? They are the result of *fitting* the learning algorithm (linear regression) to the data. Once we have them defined, we can then say we have the predictive algorithm where, if we were to have a new sample ($x_i$) with age information, we could find out the predicted glucose ($y_i'$).

As you will see later, the way this *fitting* works, is by finding the parameters \beta\_0 and \beta\_1 which make the best fit line.And this is calculated through *optimization* of the *loss/cost function* through either *ordinary least squares (OLS)* or *gradient descent* (do not worry about this just now!)

More specifically, when you use the function below `linear_reg()` or any other function that fits a linear model in R or Python, \beta\_0 and \beta\_1 or \beta\_n (depending on the amount of features you are including in your model) are being calculated using *OLS*

```{r}

lm_model <- linear_reg() %>% 
            set_engine('lm') %>% # adds lm implementation of linear regression
            set_mode('regression')


ls_fit <- lm_model %>% 
          fit(glucose ~ age, DataSmaller)


```

```{ls_fit[["fit"]][["coefficients"]]}
```

```{}
```

If we include this coefficients and draw the line ourselves, we will obtain the same thing as the `geom_smooth` above.

```{r}

#| fig-align: 'center'
#| fig-width: 6
#| fig-height: 4

ggplot(DataSmaller, aes(x = age, y = glucose)) + 
geom_point() + 
labs(x = "Age (Feature)", y = "Glucose (Label)") +  
geom_abline(intercept = ls_fit[["fit"]][["coefficients"]][1], slope = ls_fit[["fit"]][["coefficients"]][2], color="red",
               linetype="dashed", size=1.5)

```

The predictive model then (with fitted parameters) is:

$$
glucose = 73.68 + 1.33*age
$$ If we now have a new sample coming with $age = 45$, what is the predicted glucose? Can you include it in the plot?

```{r}

age_newsample<- 45
predicted_glucose <-  73.68 + 1.33*age_newsample
print(predicted_glucose)

```

This is the same as doing:

```{r}
predict(ls_fit, new_data = data.frame(age = 45)) # we had to create a new dataframe with one value of age for this to work, as it normally expects more than 1 smple, and more features!
```

```{r}
ggplot(DataSmaller, aes(x = age, y = glucose)) + 
geom_point() + 
geom_point(x = age_newsample,y = predicted_glucose, colour = "grey", size = 5 ) +
labs(x = "Age (Feature)", y = "Glucose (Label)") +  
geom_abline(intercept = coef(ls_fit)[1], slope = coef(ls_fit)[2], color="red",
               linetype="dashed", size=1.5)

```

Question!! If my new sample is $age = 10$, can I still predict glucose, even though none of my data used to create the model had a sample with age = 10?

```{r}

age_newsample<- 10
predicted_glucose <-  73.68 + 1.33*age_newsample
print(predicted_glucose)


ggplot(DataSmaller, aes(x = age, y = glucose)) + 
geom_point() + 
geom_point(x = age_newsample,y = predicted_glucose, colour = "grey", size = 5 ) +
labs(x = "Age (Feature)", y = "Glucose (Label)") +  
geom_abline(intercept = coef(ls_fit)[1], slope = coef(ls_fit)[2], color="red",
               linetype="dashed", size=1.5) + 
  xlim(0, 100)


```

Yes we can! Another thing is how confident we are about the predicted value, but we will discuss that later on.The model is an infinite line, mapping age to glucose (so we could even map negative values, or very high age values (e.g 1000) and obtain a predicted glucose - but it would make no sense).

### EXERCISE: Understand the role of the dataset in predictive modelling

Lets fit the decision tree learning algorithm in different number of training points (you will learn more about how decision trees work next Tuesday, but as you have seen in the previous classes, learning algorithms are already coded for you as functions, so you can apply them already)

```{r}

dt_model <- decision_tree() %>% 
            set_engine('rpart') %>% # check out what the rpart library/engine is about
            set_mode('regression')


dt_fit <- dt_model %>% 
          fit(glucose ~ age, DataSmaller)



```

Check out what the fitted model consist of, and print out the information the model encodes (open ls_fit too and see)

```{r}
#View(dt_fit)
dt_fit[["fit"]]
```

Now lets plot the model to get a better idea of what it is doing ( this would be the same as priniting out the intercept and coefficients for a linear regression algorithm )

```{r}

library(rpart.plot)
# Plot the decision tree
rpart.plot(dt_fit[["fit"]], type = 3, fallen.leaves = TRUE, box.palette = "Blues", shadow.col = "gray")


```



```{r}
#Get the predicted glucose values 
DataSmaller$predicted_glucose <- predict(dt_fit, new_data = DataSmaller)$.pred
```

This we are now introducing the dataset that we fitted our learning algorithm on, as if they were new samples we want to have glucose predicted on. Do you see how the real value and the predicted one differ? The closer we can get them to be the better model we will have!


**STOP HERE AND MAKE SURE YOU HAVE UNDERSTOOD THIS CONCEPT**

Do these plots help understand?

```{r}

 ggplot(DataSmaller, aes(x = age, y = glucose)) +
    geom_point(size = 3, color = "blue", alpha = 0.7) +  # Larger, colored points with some transparency
    geom_step( aes(y = predicted_glucose), colour = "darkred", size = 2) + 
    labs(
      x = "Age (years)",
      y = "Glucose Level (mmol/L)") +
    theme_minimal(base_size = 18) +  # Use a clean theme with larger base font size
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),  # Center and bold title
      plot.subtitle = element_text(hjust = 0.5),              # Center subtitle
      axis.title = element_text(face = "bold"),               # Bold axis titles for readability
      panel.grid.major = element_line(color = "grey85"),      # Lighten grid for subtlety
      panel.grid.minor = element_blank()                      # Remove minor grid lines for clarity
    ) 


```

```{r}
ggplot(DataSmaller, aes(x = age, y = glucose)) +
    geom_point(size = 3, color = "blue", alpha = 0.7) +  # Larger, colored points with some transparency
    geom_step( aes(y = predicted_glucose), colour = "darkred", size = 2) +  # Step function for
    geom_point(aes(x = age, y = predicted_glucose), size = 2, color = "grey", alpha = 0.8) + 
    labs(
      x = "Age (years)",
      y = "Glucose Level (mmol/L)") +
    theme_minimal(base_size = 18) +  # Use a clean theme with larger base font size
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),  # Center and bold title
      plot.subtitle = element_text(hjust = 0.5),              # Center subtitle
      axis.title = element_text(face = "bold"),               # Bold axis titles for readability
      panel.grid.major = element_line(color = "grey85"),      # Lighten grid for subtlety
      panel.grid.minor = element_blank()                      # Remove minor grid lines for clarity
    ) 

```

**DISCUSS THIS WITH YOUR PARTNER AND EXPLAIN IT TO EACH OTHER**

------------------------------------------------------------------------

Now lets fit the decision tree algorithm in an even smaller set of points:

```{r}
DataEvenSmaller <- DataSmaller[1:30, ]
dim(DataEvenSmaller)
```

-   

    1)  What is the model created?

-   

    2)  How does it differ from the previous?

```{r}

```

```         
3)  Do the same with the linear regression model. What changes?
```
